---
title: "lab_08"
author: "derek willis"
date: "2023-04-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab. We'll be making some charts, working with dates and retrieving Census data.


```{r}
#load libraries
library(tidyverse)
library(lubridate)
library(tidycensus)
```

## Load and modify data

**Task** Create a codeblock and load the following data from the data folder:
**Task** Run the following code to load packages.

```{r}
#load data
prince_georges_2022_overdoses <- read_csv("data/prince_georges_2022_overdoses.csv")
View(prince_georges_2022_overdoses)
```

* Prince George's County 911 Overdose Calls

You will need to create columns for the date, week and month based on the existing `datetime` column.

```{r}
#create columns for date, week, month
pg_2022_od <- prince_georges_2022_overdoses |> 
  mutate(date = date(datetime), 
        week = week(datetime), 
        month = month(datetime, label=TRUE))

```


## Questions 

**Q1.** Which month saw the greatest percentage of total calls? Create a dataframe that calculates the percentage of all calls that each month's calls represents. Make a bar chart from that dataframe. Your bar chart must have:

* A clear title that states the main idea/finding
* Good labels for the x & y axis and a caption for the source, which is Prince George's County EMS.
* Readable bars - the values shouldn't be overlapping

Compare the results here to those from the pre_lab_08 bar chart - are there major differences in the months with the highest and lowest figures? Describe that below.

**A1.** The month with the highest percentage of total calls is December. I have arranged the bar graph according to months with the highest percentage to the month with the lowest. It is arranged as such because it would be easier for me to answer this question. If it were for publication, I may have left it in chronological order as it will be easier for the reader to see it.

Comparing these results with the results in pre lab, the month with the highest percentage of number of calls in PG is December and the lowest is in January. However, the month with the highest number of calls in Baltimore is August and the lowest is February. I guess the difference is that the highest percentage of calls in PG County is during festive/holiday months, except for July which is a warm month. Highest number of calls in Baltimore is in August, which is a warmer month. 

I guess it makes sense that PG County has its highest percentage calls in December as it is in the holiday season. July, the second highest MAY BE tied to 4th of July and summer holidays. November is Thanksgiving and maybe Veteran's Day.

```{r}
pg_2022_od_pct <- pg_2022_od |> 
 group_by(month) |> 
 summarise (
   calls_month = n()) |> 
mutate(
    pct_calls_month = (calls_month/sum(calls_month)) * 100
  ) |> 
arrange(desc(pct_calls_month))
```

```{r}
pg_2022_od_pct |>
  ggplot() +
  geom_bar(aes(x=reorder(month, pct_calls_month), weight= pct_calls_month)) +
  coord_flip() +
  theme_minimal() +
  labs(
    title="Highest percentage of 911 calls are holiday months",
    x = "month",
    y = "percentage of 911 calls",
    caption = "source: Prince George's County EMS"
  )

```


**Q2.** Let's visualize this data on a weekly basis using a line chart. As in Q1, generate a dataframe with the total number of calls for each week, and then create a line chart to show the distribution of calls over time. Your line chart must have:

* A clear title that states the main idea/finding
* Good labels for the x & y axis and a caption for the source, which is Prince George's County EMS.
* Readable labels

Describe the pattern of results; you may want to look at the data to dig into particular outliers.

**A2.** 
So the weeks with the highest total calls are in week 28 which is the week of July 11 to July 17. It may be a week off but it is quite close to July 4th. The next highest week is week 51, which is the week leading up to Christmas. The third and forth week with the highest calls is week 48 and 47, which is the week around Thanksgiving. The fifth week with the highest number of calls is week 50 which is leading up to Christmas as well. These patterns are reflected on the line graph. 

One outlier in the pattern are the two peaks between week 5 and week 10. Around week 5 is the end of January and beginning of February. My guess for this may be due to it being the coldest part of the year, with a low number of sunlight hours. The other peak in week 10 would be around 7-13 March. This could be leading close to Easter, Spring Break and St. Patrick's Day. These are just guesses.

Another interesting point on the line graph is the major drop at the end of the graph which would be the end of the year. I'm sure the dip does not correspond to data as I think there should be a high number of calls leading to new year's and new year's eve. 

```{r}
pg_2022_od_week <- pg_2022_od |> 
  group_by(week) |>
  summarise (
   calls_week = n()) |> 
arrange(desc(calls_week))
```

```{r}
pg_2022_od_week |>
  ggplot() +
  geom_line(aes(x=week, y=calls_week)) +
  theme_minimal() +
  scale_x_continuous(breaks=seq(0, 53, by=5)) +
  labs(
    title="Number of 911 calls per week",
    x = "week",
    y = "total calls",
    caption = "source: Prince George's County EMS"
  )
```
**Q3.**  A scatterplot is a type of chart that helps us see relationships between two variables. One variable goes on the x axis, the other on the y axis.  For each row/observation in our data, a scatterplot puts a circle (or a "point") where the two variables intersect on a grid. 

Statisticians use scatterplots to show graphically whether one variable is correlated -- related, in a statistical sense -- with another variable.  A classic example is the [relationship between ice cream sales and temperature](https://www.mathsisfun.com/data/scatter-xy-plots.html). The scatterplot below -- press play to load the image -- shows that relationship, that an increase in temperature is associated with an increase in ice cream sales. When it's 12C, sales are 200 dollars, and when it's hotter, 25C, sales are 600 dollars.

```{r}
knitr::include_graphics("https://www.mathsisfun.com/data/images/scatter-ice-cream1.svg")
```

We're going to use a scatterplot a little differently, to get a visual sense of two key variables: 

Our question is: does the median income in a zip code have any relationship to the number of overdose 911 calls in that zip code?

To answer this question, do the following:

1. Generate a dataframe with the number of 911 calls for each zip code.
2. Get data from the Census Bureau showing median household income for Maryland zip codes.
3. Join those two dataframes on their zip code columns, starting with the 911 calls dataframe.
4. Make a scatterplot showing the total calls and median income. I didn't show you how to do this, so look it up! Googling "ggplot scatterplot" is a good start.
5. Give it an appropriate title, source, and x and y axis titles.
6. Add a label for each point that shows the zip code using geom_text() - see some examples of its use at https://ggplot2.tidyverse.org/reference/geom_text.html#ref-examples. Try to make the names as easy to read as possible by avoiding overlap.
7. In the answer space below, describe what you see and answer the questions posed above. In a general sense, what do you think this means? Feel free to consider the actual raw values: how would you report out the main point(s) of this chart?

Our question is: does the median income in a zip code have any relationship to the number of overdose 911 calls in that zip code?
**A3.** From the scatterplot below, it shows that there are more number of phone calls in zip codes with lower median incomes. There are more number of phone calls in zip codes with an estimated median income in the range of $65,000 to $80,000. The zip code with the highest number of calls according to the scatterplot is 20743, which is in Capitol Heights in PG County. According to the raw data, the zip code with highest number of calls is 20745, which is in Oxon Hill, near Alexandria in PG County. According to the raw data, the median estimate income for 20745 is $65,845 and zip code 20743 is $65,152. 

Based on this, it looks like a few of the zipcodes within the $65,000 to $80,000 median income range have the biggest range in number of phone calls. According to the Maryland map with zipcode labels, it seems that the zipcodes within that range are around South PG County. [source= https://www.unitedstateszipcodes.org/20745/]

I would like to look into the demographics and the reason why there is such a vast difference in number of phone calls for the zip codes in the median income range of $65,000 to $80,000. Looking into those demographics, I want to check which of these zipcodes have universities in them. 
```{r}
library(ggplot2)
```

```{r}
pg_calls_zipcode <- pg_2022_od |> 
  group_by(zipcode) |> 
  summarise (
   calls_week = n()) 

pg_calls_zipcode <- pg_calls_zipcode |> 
  mutate(zipcode = as.character(zipcode))
```

```{r}
acs5 <- load_variables(2021, "acs5", cache = TRUE) 
view(acs5)
```

```{r}
md_acs_census <- get_acs(geography = "zcta",
              variables = c(medincome = "B19013_001"),
              state = "MD",
              year = 2019)

pg_acs_zip <- pg_calls_zipcode |> left_join(md_acs_census, by = c('zipcode'='GEOID')) 

```
```{r}
pg_acs_zip |> 
ggplot() +
  geom_point(aes(x=estimate, y=calls_week)) +
  theme_minimal() +
  labs(
    title="Higher number of OD calls made in zipcodes with lower median income",
    x = "estimated household median income",
    y = "OD calls per week",
    caption = "source: Prince George's County EMS, Each point represents a zip code"
  ) +
  geom_text(aes(x=estimate, y=calls_week+5, label=zipcode), check_overlap=TRUE) 

```
```