---
title: "lab_03"
author: "Derek Willis"
date: "2023-09-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse

# Load janitor
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab: the Maryland payments to vendors and Maryland grants and loans and, using janitor, change the columns so that they are lower-cased and any spaces are replaced with underscores.

```{r}
payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=373564)

# cleaning function
cleaned_md_grants_loans <- md_grants_loans |>
  clean_names()
# display the cleaned dataset
cleaned_md_grants_loans

#cleaning function
cleaned_md_payments_data <- payments
  clean_names()
#display the cleaned dataset
cleaned_md_payments_data
```
Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed. Write code to find rows where the vendor name contains the county's school system, starting with a broad search and then narrowing it down. When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find? Look at & search the records in the dataframe while you construct your code.

A1:15 records

```{r}
pg_schools <- cleaned_md_payments_data |> 
   filter(str_detect(vendor_name, "BOARD OF EDUCATION OF PRINCE GEORGE"))


```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2:The fiscal year 2022 had the largest amount at $1,578,857,325. It shows that payments to PG schools have increased in the last 10 years. Another interesting thing is that the fiscal years are 2012, 2016 and 2022 and I wonder why it is during these years.

```{r}
pg_schools |> 
  group_by(fiscal_year) |> 
    summarise(total_amount=sum(amount)) |>   
    arrange(desc(total_amount))
    
```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received. Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3.The cleaning required would be the distinct function to case_when the vendor_names. I realised that some vendor names are similar and could be the same but have zip codes where one or two numbers are different.

Three examples of vendors names that are similar: BRAWNER BUILDERS INC, WHITING TURNER CONTRACTING CO and BARTON MALOW COMPANY.

I would use the case_when function to add the companies similar in name into one name. For example, WHITING TURNER CONTRACTING CO and WHITING TURNER CONTRACTING COMPANY to WHITING TURNER CO.

```{r}
cleaned_md_payments_data |> 
  filter(
    str_detect(agency_name, "CAPITAL PROJECTS")
  ) |> 
group_by(vendor_name, vendor_zip) |> 
  summarize(total_amount = sum(amount)) |>
    arrange(desc(total_amount))
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which columns contain the information you want, your code should standardize the data needed to find just the grants for stem cell research in a single filter. Then create a total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4:It shows that the total amount of stem cell research funding has decreased from 2009 to 2018, with the funding being half the total amount it was in 2009.

```{r}
cleaned_md_grants_loans |> 
   filter(str_detect(description, "Stem Cell Research Grant")) |> 
   group_by(fiscal_year) |> 
   summarize(total_amount = sum(amount)) |>
    arrange(desc(total_amount))
  

```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5: The number of grantees are 18. I would use the mutate(str_to_upper) function to standardise the names of the grantees. I would then use the case_when function to put grantees under University of Maryland if they have similar combinations.I think Haystack solutions is interesting since it is a cyber security company. I understand that cyber security is important for a university, especially with research and projects. The other thing I find interesting is that 4 companies had total amounts of $100,000.00. The companies are GripBoost, Maryland Energy & Sensor Technology LLC, MedSense LLC, PIXELLIGENT TECHNOLOGIES, LLC and it would be interesting to know that was the amount suggested as a minimum or maximum to the companies.

```{r} 
cleaned_md_grants_loans |> 
  filter(str_detect(zip_code, "20742")) |> 
  group_by(grantee) |> 
  summarize(total_amount = sum(amount)) |>
    arrange(desc(total_amount))
  
```

