---
title: "Graduate Assignment: Motor Vehicle Theft Analysis"
author: "Daranee Balachandar"
date: "2023-11-26"
output: html_notebook
---
#Maryland Property Crime and Motor Vehicle Theft Analysis

For this data analyisis project, I will be working with the Violent Crime & Property Crime by Municipality: 2000 to Present Data (https://opendata.maryland.gov/Public-Safety/Violent-Crime-Property-Crime-by-Municipality-2000-/2p5g-xrcb) to look into how prevalent MV theft is in Maryland, the demographics and geographic of such thefts and if there is any behavioural change in Maryland citizens due to the crime.

## Data set up

Load libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
# install.packages("ggrepel")
# install.packages('ggthemes')
library(tidyverse)
library(lubridate)
library(janitor)
library(ggthemes)
library(tidycensus)
library(ggplot2)
library(ggrepel)
```

Load Data

```{r}
Violent_Crime_Property_Crime_by_Municipality_2000_to_Present <- read_csv("data/Violent_Crime___Property_Crime_by_Municipality__2000_to_Present.csv")
```

Story Ideas
1. Top 5 hotspots for MV thefts in MD and if there is a common identifier of demographic in these 5 spots. 
2. Have these MV theft stories decreased the number of car sales of Kias and Hyundais in MD.
3. Are there any other crimes in MD that should have more attention other than this

