---
title: "lab_04"
author: "Derek Willis"
date: "2023-10-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

You'll need to load three packages for this: the tidyverse, lubridate and janitor.

**Task** load these three packages.

```{r}
# Turn off scientific notation
options(scipen=999)
library(tidyverse)
library(lubridate)
library(refinr)
# load your packages

```

For this lab, we want to investigate spending by Maryland state and local candidates during the previous election. For example, we could ask how much money have candidates and committees spent on advertising on Facebook? Which vendors received the most money for media expenses? We have the data, but it's messy - names of payees are spelled differently and capitalized differently - and we need to clean it up. We'll use a combination of RStudio and OpenRefine to do that.

The basic process here is to start in RStudio, export data so that we can use OpenRefine to clean it up, and then bring it back into RStudio to finish our analysis.

## Load Data

You'll need to load one data set, a CSV file of Maryland campaign expenditures located in the data folder called "maryland_expenses.csv"

**Task** Create a codeblock below, then read the data in and assign it to an appropriate variable name. You'll want to clean up the column names and make sure that any date columns are actually date datatypes.

```{r}
# Load the Maryland expenditures data table
maryland_expenses <- read_csv("data/maryland_expenses.csv")

# Display it
maryland_expenses

install.packages('janitor')

library(janitor)

# cleaning function
cleaned_maryland_expenses <- maryland_expenses |>
  clean_names()

# display the cleaned dataset
cleaned_maryland_expenses

cleaned_maryland_expenses <- cleaned_maryland_expenses |> mutate(expenditure_date=mdy(expenditure_date))

head(cleaned_maryland_expenses)
```

## Answer questions

**Q1.** You've been assigned to write about Maryland political committees' use of Facebook advertising. There's a category column in the data that has values for different kinds of spending. One of them is "Media". You'll need to complete the following steps:

   1. Create a dataframe of Media expenses where the name of the payee is not NA. 
   So I used ChatGPT for this and it told me to use this code filter(!is.na(Value) to remove the NA. Apparently ! means to remove or without.
   2. Then, using `write_csv`, make a CSV file of your media expenses in your data folder. `write_csv()` is similar to `read_csv()` but you provide it with the dataframe you're writing and the file path.
   3. Using OpenRefine, create a project using that CSV file, create a copy of the `payee_name` column called `payee_clean` and then standardize it, focusing on the payees that appear most often. Pay particular attention to variations of Facebook, bearing in mind that it changed its name in 2021: <https://about.fb.com/news/2021/10/facebook-company-is-now-meta/>), but also clean up variations of the most common payees.
   4. When you are done standardizing `payee_clean`, export it as a new CSV file to your data folder, *giving it a different name* and read it back into RStudio as a new dataframe.
   5. Using that `payee_clean` column in your new dataframe, calculate the total amount of Facebook spending by each committee and present the results showing the committee that spent the most first.

**A1.** It is the Jain Ashwani Friends of committee with the total amount of $62,357.38. I think it's actually called Friends of Ashwani Jain.

```{r}
media_expenses <- cleaned_maryland_expenses |> 
   filter(str_detect(expense_category, "Media")) |> 
   filter(!is.na(payee_name))

write_csv(media_expenses, "data/media_expenses.csv")

media_expenses_md <-read_csv("data/media_expenses_md.csv") 

media_expenses_md |> 
  filter(str_detect(payee_clean, "Facebook")) |> 
  group_by(committee_name,payee_clean) |> 
  summarise(total_amount = sum(amount)) |>
  arrange(desc(total_amount))
   
```

**Q2.** Let's do some research on the top committees from question 1. Look up the top 10 spenders: if they are candidate committees, what offices were they running for? If they were not candidates, what issues do they seem to care about? Write some sentences answering those questions, then go the Facebook ad library ([https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data\\[direction\\]=desc&sort_data\\[mode\\]=relevancy_monthly_grouped&media_type=all](https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data\%5Bdirection\%5D=desc&sort_data\%5Bmode\%5D=relevancy_monthly_grouped&media_type=all){.uri}) and look for ads from the top spender. What do you notice about the the information provided about those ads?

**A2.** So 8 out of the top 10 spenders are candidate committees. Most of them are running for Governor of Maryland. In the 8 of them, 4 of them are running for Governor of Maryland, for example, Jain Ashwani Friends of, Franchot Anderson-Walker for Maryland Slate, Franchot Peter Friends Of, Baron Jon (Jonathan) for Maryland, King and Siri for Maryland Slate. Please note that  Franchot Anderson-Walker for Maryland Slate, Franchot Peter Friends Of are two different committees but it is for Peter Franchot, who is running for Governor in Maryland. So, the top spender is Friends of Ashwani Jain. I noticed that most of the ads were put out in July 2022 and the amount spent for ads are between less than $100 to more than $1,000. And the estimated audience size are almost mostly more than 1 million audience. I also noticed that quite a few of the ads were removed.

**Q3.** Given the list of top spenders in A1 and the primary results for governor (<https://elections.maryland.gov/elections/2022/primary_results/gen_results_2022_1.html>), who is missing from that Q1 list? Did those campaigns run Facebook ads? Who are the biggest recipients of media expenses by the Democratic primary winner for governor (HINT: that person has two committees, the second one with his running mate)?

**A3.** Multiple republican candidates are missing from the Q1 list. And most importantly, the democratic primary winner for governor Wes Moore and his running mate Aruna Miller. Yes, the Wes Moore campaign did run facebook ads. It is SKDK, a strategic communications organisation at $6,172,801.71, then it is Moore Miller for Maryland State at $4,867,000.00. So, I guess Wes Moore campaign committee basically pays SKDK to pay for Facebook ads.

```{r} 
cleaned_maryland_expenses |> 
  filter(str_detect(committee_name, "Moore")) |> 
  group_by(payee_name) |> 
  summarise(total_amount = sum(amount)) |>
  arrange(desc(total_amount))
```

**Q4.** The media expense data you cleaned also has a standardized purpose column. Write code to generate a list of the committees with the total amount of spending on online advertising. How does this list compare to the one from Q1 about top Facebook spenders? Of the top 5 committees who spent the most on Facebook ads from Q1, are there any that appear to have spent *all* of their online advertising dollars on Facebook?

**A4.** Interestingly, comparing the top 10 spenders in Q1 and the top 10 in Q4, Casa in Action PAC seems to be the committee that pops up in both lists. But, I'd like to also note that Baron-Williams Slate for Maryland did pop up in the Q1 list as Baron Jon (Jonathan) for Maryland. Of the top 5 committees, Casa in Action PAC appears in the online advertising list and they spent $20,647.83 out of $536,496.16 in Facebook ads. So, no there aren't any from the top 5 committee who has spent all their oline advertising dollars on Facebook.

```{r}
cleaned_maryland_expenses |> 
  filter(str_detect(expense_purpose, "Online Advertising")) |> 
  group_by(committee_name) |> 
   summarise(total_amount = sum(amount)) |>
  arrange(desc(total_amount))
```

**Q5** Your editor asks you to summarize your findings about spending on Facebook by Maryland political candidates. How much do we know and what don't we know? What's the best story idea you've found here?

**A5** Most Democrat campaigns or political candidates tend spend on Facebook ads. Just a question and an area to look into, where do Republican political candidates and campaigns put their ad campaign in if it is not in Facebook? Why not Facebook? Is it Twitter or X?
